Code used to generate paper results : Database comparison and CT characteristics

### 2024-05-23 Paper_code-CT_analysis.Rmd
Added some heatmaps

@@

@@

@@

  ## Use gene names instead of ENSEMBL IDs
  mat <- SummarizedExperiment::assay(TCGA)
  rownames(mat) <- rowData(TCGA)$external_gene_name
  mat <- mat[CT_genes$external_gene_name,]
  mat <- log1p(mat)

@@

  ## Use gene names instead of ENSEMBL IDs
  mat <- SummarizedExperiment::assay(CCLE)
  rownames(mat) <- rowData(CCLE)$external_gene_name
  mat <- mat[CT_genes$external_gene_name,]
  mat <- log1p(mat)

@@

>>>>

all_genes_types <- all_genes %>% 
  dplyr::rename(gene = external_gene_name) %>% 
  mutate(Regulation = case_when(regulated_by_methylation ~ "Methylation",
                                !regulated_by_methylation ~ 
                                  "Not methylation")) %>%
  mutate(Chr = case_when(X_linked ~ "X-linked",
                         !X_linked ~ "not X")) %>% 
  mutate(chr_met = case_when(regulated_by_methylation & X_linked ~ "X_met",
                             !regulated_by_methylation & X_linked ~ "X_not_met",
                             regulated_by_methylation & !X_linked ~ "not_X_met",
                             !regulated_by_methylation & !X_linked ~ 
                               "not_X_not_met")) %>% 
  mutate(specificity = case_when(GTEX_category == "testis_specific" ~ 
                                   "Testis specific",
                                 GTEX_category == "testis_preferential" ~ 
                                   "Testis preferential",
                                 GTEX_category == "lowly_expressed" ~ 
                                   "Undetectable in GTEX")) %>% 
  mutate(specificity = factor(specificity,
                              levels = c("Testis specific",  
                                         "Undetectable in GTEX", 
                                         "Testis preferential"))) %>% 
  mutate(chr_met = factor(chr_met, 
                          levels = c("X_met", "not_X_met", "X_not_met", 
                                     "not_X_not_met")))

@@

# Genes that were selected with GTEx but leaky so removed (ubiquitary expression)
all_genes %>% 
  filter(GTEX_category %in% c("testis_specific", "testis_preferential")) %>% 
  filter(TCGA_category == "leaky" | CCLE_category == "leaky") %>% 
  dim()

@@

  gene_type <- all_genes_types %>% 
    filter(gene %in% genes)

@@

  ## Use gene names instead of ENSEMBL IDs
  mat <- SummarizedExperiment::assay(TCGA)
  rownames(mat) <- rowData(TCGA)$external_gene_name
  mat <- mat[rownames(mat) %in% all_genes$external_gene_name,]
  mat <- log1p(mat)

  gene_type <- all_genes_types %>% 
    filter(gene %in% genes)

@@

  ## Use gene names instead of ENSEMBL IDs
  mat <- SummarizedExperiment::assay(CCLE)
  rownames(mat) <- rowData(CCLE)$external_gene_name
  mat <- mat[rownames(mat) %in% all_genes$external_gene_name, ]
  mat <- log1p(mat)

  gene_type <- all_genes_types %>% 
    filter(gene %in% genes)

@@

Counter examples

```{r counter_examples}
genes_not_testis <- all_genes %>% 
  filter(GTEX_category != "lowly_expressed") %>% 
  filter(testis_specificity == "not_testis_specific") %>% 
  pull(external_gene_name) %>% 
  sample(2)

genes_not_TCGA_activated <- all_genes %>% 
  filter(GTEX_category != "lowly_expressed") %>% 
  filter(testis_specificity != "not_testis_specific") %>% 
  filter(!TCGA_category %in% c("activated", "multimapping_issue")) %>% 
  pull(external_gene_name) %>% 
  sample(1)

genes_not_ccle_activated <- all_genes %>% 
  filter(GTEX_category != "lowly_expressed") %>% 
  filter(testis_specificity != "not_testis_specific") %>% 
  filter(CCLE_category %in% c("activated")) %>% 
  pull(external_gene_name) %>% 
  sample(1)

my_genes_counter <- c(genes_not_testis, genes_not_TCGA_activated, 
                      genes_not_ccle_activated)

ht_TCGA_counter <- print_TCGA_heatmap(my_genes_counter, 
                              fontsize = 4, 
                              downsample_to_n = 50, 
                              h_width = 3,
                              h_height = 8,
                              tum_type = c("TCGA-SKCM", "TCGA-LUAD",
                                           "TCGA-HNSC"),
                              clust_method = "ward.D",
                              split_rows_by = "specificity")

ht_CCLE_counter <- print_CCLE_heatmap(my_genes_counter, 
                              downsample_to_n = 50, 
                              tum_type = c("Skin", "Lung",
                                           "Head_and_Neck"),
                              h_width = 3,
                              h_height = 8,
                              split_rows_by = "specificity")

ht_GTEX_counter <- print_GTEX_heatmap(my_genes_counter, 
                              row_fontsize = 8,
                              column_fontsize = 5,
                              h_width = 5,
                              h_height = 8,
                             split_rows_by = "specificity")
h_list_counter<- ht_GTEX_counter + ht_CCLE_counter + ht_TCGA_counter
draw(h_list_counter, main_heatmap = "TCGA (logTPM)", ht_gap = unit(c(0.3, 0.01),
                                                                   "cm"),
     merge_legend = TRUE)

my_genes_leaky <-  all_genes %>% 
  filter(GTEX_category %in% c("testis_specific", "testis_preferential")) %>% 
  filter(TCGA_category == "leaky" | CCLE_category == "leaky") %>% 
   pull(external_gene_name)

ht_TCGA_leaky <- print_TCGA_heatmap(my_genes_leaky, 
                              fontsize = 4, 
                              downsample_to_n = 50, 
                              h_width = 3,
                              h_height = 8,
                              tum_type = c("TCGA-SKCM", "TCGA-LUAD",
                                           "TCGA-HNSC"),
                              clust_method = "ward.D",
                              split_rows_by = "specificity")

ht_CCLE_leaky <- print_CCLE_heatmap(my_genes_leaky, 
                              downsample_to_n = 50, 
                              tum_type = c("Skin", "Lung",
                                           "Head_and_Neck"),
                              h_width = 3,
                              h_height = 8,
                              split_rows_by = "specificity")

ht_GTEX_leaky <- print_GTEX_heatmap(my_genes_leaky, 
                              row_fontsize = 8,
                              column_fontsize = 5,
                              h_width = 5,
                              h_height = 8,
                             split_rows_by = "specificity")
h_list_leaky<- ht_GTEX_leaky + ht_CCLE_leaky + ht_TCGA_leaky
draw(h_list_leaky, main_heatmap = "TCGA (logTPM)", ht_gap = unit(c(0.3, 0.01), "cm"),
     merge_legend = TRUE)
```
